\section{Introduction}

\subsection{General Overview}

Evidence based decision making is widely presented as the best way to design public health policies at national and local levels \citep{abou-zahr_health_2005,shibuya_health_2005,bambas_nolen_strengthening_2005,mutemwa_hmis_2006,boerma_public_2013}.
The ability of statistics to offer rational basis to make decisions \citep{desrosieres_politique_1993,porter_trust_1996},
allied with progress in data collection and analysis techniques engender enthusiasm and hope in the ability of governments and other actors to design and implement analytical tools to improve the political and administrative processes.

If this principle seems to be set, what constitutes relevant evidence in an administrative context, and how it should be generated is less clear and depends on local contexts and traditions. Bergeron and Cassel note that knowledge and expertise of public health deciders are "a network involving many different actors, structures and tools, concepts and spatial and institutional arrangements" \citep{bergeron_savoirs_2014}. Producing knowledge for public health decision should thus relies on localized and contextual approaches. In the world of Global Health Metrics meanwhile, the local and contextual dimension of statistical systems can sometimes be lost in an undue generalization. This is in part due to the fact that the way people in charge of statistical systems and policy design in developing countries have been influenced by a mindset that favors aggregated high level evidence over local contextualized knowledge, and standardization over adaptation.

Indeed, in most rich countries, a history of public statistics development resulted in the emergence and stabilization of well defined data sources, of methods and of roles along which the generation of public statistics are produced. In these settings, strong public health information systems really are the result of local equilibriums and of adaptations to local statistical cultures, started by individuals making ad-hoc use of different data sources and inventing and standardizing methods on the run \citep{lecuyer_medecins_1987}.

On the other hand, in sub-Saharan Africa, evidence generation for public health policy makers has a very different tradition. The weakness of statistical systems in sub-Saharan Africa has been well described and documented, and have their roots in a history of colonial heritage and post-colonial developmentalist politics \citep{jerven_poor_2013}. As a result, most sub-Saharan countries can't rely on a robust vital statistics system to plan population targeted intervention. National Health Information Systems are often considered weak \citep{abou-zahr_better_2010,kiberu_strengthening_2014}, and are affected by multiple uncoordinated demands for data collection and reporting. These demands may come from local authorities or from international donors and partners, building incentives in the evidence generation process that may tend to jeopardize the accuracy of the collected information \citep{sandefur_political_2013}.

The development of statistical systems in sub-Saharan countries can indeed  be traced, in a long perspective, to the enforcement of  a specific mode of administrative control by colonial powers in the XIXth century \citep{appadurai_number_1996,cordell_couting_2010,gervais_how_2010}.
Colonial statisticians, often weakly skilled and trained \citep{kateb_gestion_1998,cordell_couting_2010}, nonetheless set the nomenclatures and conventions around how land and populations would be described and analyzed \citep{rambert_cartographie_1922,gervais_how_2010}.
Whereas European statistical systems were developed and structured by social activists in the context of the design of the first modern welfare systems \citep{desrosieres_politique_1993,desrosieres_administrator_1997}, colonial statistical systems were geared towards efficient land administration and exploitation \citep{rambert_cartographie_1922,de_martonne_cartographie_1931} with little interest on local population or analytical finesse.

After the decolonizations, this administrative mindset was completed by an expert mindset of the developmental state \citep{bonneuil_development_2000}. Developmentalist expertise transformed African societies into objects of studies and knowledge, prolonging the outwards orientation of colonial statistics. The development and the rollout of neoliberal statistical management in the 90's, and the generalization of \gls{me} for different projects \citep{desrosieres_prouver_2014} only strengthened and deepened this tendency. As the result, some characteristics of these systems continue to be a low investment in data collection, an excruciating exigency for unrealistic precision, the importation of external methods and definitions, the overarching generalization of categories for administrative simplicity. All these features are characteristic of systems that are designed for centralized administration of large domains, but are not well fit for local decision making, adaptation to local realities, and production of autochthonous knowledge of communities.

\subsection{Work hypothesis}

My main work hypothesis the top-down approach used in the design and implementation of health information systems in many developing countries is ill suited for their purpose, and is an important reason for their perceived weak performance.

By top-down I mean systems in which the purported usage of information calls for specific data collection and analytical tools. Most HMIS design recommendations for an ex-ante design of both data analysis and measure framework, from which data collection should be inferred. \citep{lippeveld_routine_2000,rhino_introducing_2003,daltilia_systeme_2005,health_metrics_network_framework_2008}. For example, \gls{me} guidelines such as those of Global Health agencies like the Global Fund exercise important drives on the organization of health information systems in sub-Saharan African countries on which they are imposing definitions, classifications and reporting methods, to provide sufficient information for the evaluation of funded programs \citep{the_global_fund_global_2014}. This asymetric relationship between international and local actors leads to a    \textit{reductio ad \gls{me}} of information systems, that are built to answer imposed norms but do not create their own logics, cultures and traditions.

The mindset behind these approaches is influenced both by the colonial administrative tradition, for which standardized statistical tools should provide a normed information, and by the developmentalist expertise tradition, along which an hyper directed data collection system is aimed at providing a unique and narrowly defined piece of knowledge. This mindset is still present in the current tendency of international donors to encourage the development of systems specifically aimed at answering their own particular information needs.


HIearchiy of eveidence

National Health Information Systems are then organized around these multiple narrowly defined needs for information, and are directed towards answering questions asked by national or international specialists, with little attention at local level uses. Meanwhile, to answer the current needs of Ministries of Health and other local public health actors, it appears essential to update this old mindset, building on recent methodological and technical developments in the field of data science.


I wnt to question this appraoch

Movement towards qurstionnaing what eveidene is useful< Cf RCT.

Heqlth MEtrcis as a field should be more interested in this low level of evidence / high technicity

My methodological approach is built on the dialogue between a critical reading of current statistical practices, and the design and implementation of innovative approaches. The critic statistical systems is often made from a postmodernist perspective, questioning the structures and processes that produce statistical knowledge. This critic  often poses a radical critic of the quantitative approach to public affairs, offering little place for incremental improvements \citep{rottenburg_world_2016}. Building on Latour's call for a critic oriented \textit{towards} its object to "offer an arena in which to gather" \citep{latour_why_2004}, my work is aimed at defining what a different approach to building information systems could be. In this dissertation, I explore ways through which a top-down approach to evidence generation in health systems can be reversed, to promote  bottom-up data analysis methods that put more emphasis on data already being collected in the developing countries, and from which relevant information is extracted in an ad hoc way.

This bottom-up approach also builds on the latest developments of modern data science that are seldom found in routine health information systems in developing countries.

\begin{description}
\item[Data Centrality] I take the available data as an object of study \textit{per se}, from which the analysis starts, and which conditions all subsequent developments. This means both adapting analysis aims to available data, and integrating the data collection processes in the models of interest. A corollary of this approach is a more complete expression of the uncertainty of the measure of these quantities, which are increasingly part of the Global Health Metrics practices \citep{murray_towards_2007}.
\item[Test and validation] Strong national statistical systems emerged from both norm and local improvisation, through trial and error \citep{lecuyer_medecins_1987,chaperon_information_1988}. Incrementally adapting current metrics and  methods through localized innovation is essential, but the conditions and methods to do this are hard to create. As I develop, test and validate methods I will develop ideas and examples on how this could be made.
\item[Bayesian Perspective] Finally, most of the methods used in my dissertation tend to favor a Bayesian perspective on data modeling. Using a Bayesian approach in administrative systems has been noted to be a promising approach\citep{fienberg_bayesian_2011,little_calibrated_2012}. Bryant and Graham also noted how Bayesian methods allow a good combination of a diversity data sources and proper handling of uncertainty \citep{bryant_bayesian_2013}. In my different projects, I will work on presenting and displaying the benefits of this Bayesian approach to use and analyse health information systems data.
\end{description}


\subsection{Aims}

My principal aim will be to define, implement and test \textit{bottom up} approaches to improve Health Information Systems data use in developing countries. This aim will be reached through three specific aims, which address different key processes in health information systems. The Health Metrics Network defined the three main processes of Health Information Systems as the indicators definition and standardization, data sources and data management (database hosting, data cleaning and data processing) \citep{health_metrics_network_framework_2008}.

I will address each of these processes with a specific aim, and will explore how methods based on available data and processes can improve the performance of Health Information Systems.

\pgfdeclarelayer{background}
\pgfdeclarelayer{foreground}
\pgfsetlayers{background,main,foreground}
\begin{center}
 \begin {figure}[ht]
        \centering
\resizebox{\linewidth} {!} {
\input{graph_process_aims.tex}
}
\caption{HIS Processes and corresponding dissertation aims}
\label{fig:process_aim}
\end{figure}
\end{center}


\paragraph{Indicators / Flexible Standards} Defining categories and metrics based on which people are going to be counted is an essential piece of the statistical work \citep{desrosieres_politique_1993}. It is an essential step in the simplification involved in the activity of measurement. The field of Global Heath relies on important taxonomies, like the International Classification of Diseases and on Metrics like the Disability Adjusted Life Years, to unify description and measurement of health across the globe, and allow comparison and benchmarking \citep{murray_towards_2007,murray_health_2008}. Meanwhile, at local level, the use of globally defined metrics may have its limits, as it does not allow adaptation to local contexts and situations. The measure of retention of HIV patients in care is an example of this. Understanding the outcome of patients after they enter care is essential to evaluating HIV care systems performance, and efforts have been made to track measure to inform strategic planning and programs evaluation   \citep{the_global_fund_global_2014}. Meanwhile, the measure used in most settings, namely the proportion of patients that are considered \gls{ltfu} after a certain time in care, is an indicator that was originally designed and used in clinical research settings. The high variation of  \gls{ltfu} rates between programs and the low specificity of this metric leads researcher to question the way Loss to Follow Up is defined and retention is measured globally \citep{chi_universal_2011,yehia_comparing_2012,grimsrud_impact_2013,forster_electronic_2008}. Using \gls{emr}, I will model how the measure of retention is affected by local contexts and data quality, and I will explore  more robust ways to measure retention in HIV care.

\paragraph{Data Sources / Data Hybridization} Unavailability of population data in a large number of countries is a well known issue \citep{mahapatra_civil_2007,mikkelsen_global_2015}. As a result, spatial distribution of populations, an essential piece of evidence to develop public health policies, is often unavailable at a policy relevant scale. To palliate these shortcomings, approaches for mapping of populations have been developed, involving the use of macro-level rasters of covariates such as land coverage or night lighting imagery \citep{linard_population_2012,stevens_disaggregating_2015}. These approaches give interesting insights into how populations are distributed, and their output can easily be used for other public health work using the GPS coordinates system. Meanwhile, in situations with very low information on populations, the results of these top-down approaches is often little more than an overlay of covariates. Additionally, their results are hard to use for public health policy planning, as they do not link population to commonly used localization conventions such as places names. My second aim is to hybridize multiple data sources on population to produce a population map of Niger linked to the lowest level of population settlement possible.

\paragraph{Data management / Data Monitoring} Data quality is an important concern of Health Information Systems professionals \citep{shrestha_data_2000} who link it directly to the ability of Information systems to provide good information \citep{mphatswe_improving_2012}. The definition of data quality is directly linked to an idea of trustworthiness, which may affect the use of certain data sources to generate evidence \citep{gimbel_assessment_2011}, and the quality of health data receives a large amount of attention even from high level actors in Global Health \citep{abou-zahr_better_2010}. The best way to ensure data quality is through routine audits of primary data \citep{ronveaux_immunization_2005}. Meanwhile, this approach is costly and may have some cost-efficiency issues. Additionally, it makes little use of historical data, which makes it vulnerable to issues in primary data. My third aim will be to develop a cost effective approach to data quality screening informed by historical data and including a notion of risk attached to data quality.

\subsection{Novelty and scientific contribution}

My dissertation is contributing to three main areas of research surrounding Health Information Systems. One is mostly investigated by the \gls{ict} and interrogates the importance of local adaptation of information systems and its impact on the definition of standards. The second domain of relevance is more linked to the Global Health field, and on works that interrogate how data collected inside health systems can be used to best inform decision making. The last domain of contribution is mainly of interest for the statistical field, and explores the use of statistical methods for surveillance in applied fields.

\subsubsection{Local Adaptation and flexible standards}

In the ICT field, defining standards for data collection systems that can be implemented at local levels but respond to national or international norms. Jørn Braa, describing the approach that presided to the design and development of the \gls{dhis2} remarks that "the top-down and all-inclusive approach to standardization [is] common among ministries and central agencies" and pleads for \textit{flexible standards} following the idea that "the individual standards must be crafted in a manner which allows the whole complex system of standards to be adaptive to the local context" \citep{braa_developing_2007}. The need for local adaptability of Information Systems is seen as a key issue of \gls{ict} in developing countries \citep{macfarlane_harmonizing_2005,walsham_research_2006,walsham_foreword:_2007,jacucci_standardization_2006} and thus some \gls{ict} solutions have been designed and explored in the forms of tools like \gls{dhis2} or standards like the Open Health Information Exchange initiative.

Aims 1 and 3 explore ways in which standards for indicators definition or for the evaluation of data quality can be adapted to specific contexts. In both situations, flexibility is introduced through objective rules : specificity and robustness of the measure of retention in the Aim 1, and cost-effectiveness in the Aim 3. The two rules are different in nature, as one is internal, and based on the characteristics of the data while the other is external, and based on operational characteristics such as the cost of supervision. This approach is innovative, and can be applied to a diversity of questions. An extension of this work could also be the definition of methods for aggregation and comparison of metrics defined and measured using flexible standards.

\subsubsection{Imperfect data Usage}


Health Information Systems data is often underused, or not used at all by its intended users\citep{health_metrics_network_framework_2008}. This underuse is often blamed on the  perceived bad quality of primary data that would make it unfit for statistical analysis
\citep{ronveaux_immunization_2005,makombe_assessing_2008,heunis_accuracy_2011,gimbel_assessment_2011,who_assessment_2011,hahn_where_2013,kihuba_assessing_2014,glele_ahanhanzo_data_2015}. Approaches to solve this issue have mainly focused on improving primary data quality to improve data use  \citep{braa_improving_2012,mutale_improving_2013,ledikwe_improving_2014,nisingizwe_toward_2014}, but some authors have also pointed to the possibility to use even imperfect routine data to answer specific public health questions\citep{gething_improving_2006,gething_information_2007,wagenaar_using_2016}. These approaches rely on the nature of routine data, which is usually highly dimensional and structured times series. This type of data enables the use of robust modeling methods at an adequate level of aggregation, to control and correct for data errors.

My dissertation contributes to this line of research in two different ways.

Aim 1 and 3 also offer improved insights in routine data quality and its impact on health indicators. Aim 1  explores a micro level modeling of data quality and its impact on the measure of patient retention. It will also define retention indicators less sensitive to data quality. Aim 3 uses expected value imputation and aberrant values detection, which is an essential aspect of understanding data quality. It will also implement solutions  to analyze and present the type of data often observed in health systems.

Aim 1 and aim 2 also examine methods to enrich data, using \gls{emr} metadata or  hybridizing different data sources. Both approaches are an extension of usual approaches to health data, and make full use of the richness of modern data collection tools, and of the wide availability of public data sources.

\subsubsection{Statistical methods for health data monitoring}

For aim 3, I am interested in three types of literature that address issues of health data monitoring:
epidemiological surveillance, health service quality monitoring and industrial \gls{spc}.

I will explore a combination of those approaches to provide actionable information on available data in a health system. Indeed, all of these approach rely on a common method, based on the theoretical definition of an expected distribution of some indicator, and then defining a test to classify the observed values as coming from an \textit{in control} or \textit{out of control} process. Using a rare and important data set, I will test and benchmark different algorithm on their ability to detect different types of \textit{out of control} situations.

In the meantime, my implementation of  quality monitoring algorithms will differ from existing literature in an essential way. In most quality monitoring applications, the quantity of interest is the success of some type of procedure, measured by an outcome, of which death is usually the least desirable. In the context of Results Based Funding, on the other hand, the measure of success is much more a measure of access to a wide range of services. The denominator of such a measure is much less clear, as it is not defined at facility level, and there is often little information on target populations. I will thus use methods recently developed in the \gls{spc} field, which aim at monitoring profiles of indicators sets, in order to detect diverging patterns from a standard situation\citep{woodall_current_2014}.  The use of Profile Monitoring for this project is an important innovation in the field of surveillance.

\subsection{Timeline}

Aims specific timelines are explained in more detail in sections \ref{timeline:aim1}, \ref{timeline:aim2} and \ref{timeline:aim3}. I will prioritize getting early results on Aim 2 to allow feedback and field testing from Bluesquare and have an opportunity to reorient my methods if needed. This will also leave time me to obtain potential additional data for Aims 1 and 3. All the aims should be finalized and papers written by the end of the first quarter of 2018, which would allow for a finalization of the dissertation by June 2018.


\begin{figure}[!ht]
	\begin{ganttchart}[vgrid ,
                    hgrid,
                    y unit chart=.6cm]{1}{18}
		\gantttitle{2017}{12}
		\gantttitle{2018}{6} \\
		\gantttitlelist{1,...,12}{1} \gantttitlelist{1,...,6}{1}\\

        \ganttset{bar/.append style={draw=red!40 , fill=red!40},
                    group/.append style={draw=red, fill=red}}
		\ganttgroup{Aim 1}{1}{16} \\
        \ganttbar{Data Extraction}{1}{4} \\
		\ganttbar{Modeling and Model estimation}{5}{6} \\
		\ganttbar{Cohort Simulation}{7}{9} \\
		\ganttmilestone{Sharing Cohort simulation results}{9} \\
		\ganttbar{Data Quality Impact}{10}{10} \\
		\ganttbar{Data Maturity}{11}{12} \\
		\ganttbar{Robust Measures of retention}{13}{14} \\
		\ganttmilestone{Sharing Final Results}{14} \\
		\ganttbar{Paper Writing}{15}{16} \\
		\ganttmilestone{Paper Submission}{16} \\

        \ganttset{bar/.append style={draw=green!40 , fill=green!40},
                    group/.append style={draw=green, fill=green}}
        \ganttgroup{Aim 2}{1}{16} \\
        \ganttbar{Data Extraction}{1}{3} \\
		\ganttbar{Name Matching}{1}{6} \\
		\ganttbar{Locality Mapping}{7}{9} \\
		\ganttmilestone{First complete map}{9} \\
		\ganttbar{Population Estimation}{10}{14} \\
		\ganttmilestone{Sharing Final Results}{14} \\
		\ganttbar{Paper Writing}{15}{16} \\
		\ganttbar{Dashboard Design}{15}{16} \\
		\ganttmilestone{Paper Submission}{16}\\

        \ganttset{bar/.append style={draw=blue!40 , fill=blue!40},
                    group/.append style={draw=blue, fill=blue}}
        \ganttgroup{Aim 3}{1}{14} \\
        \ganttbar{Data Extraction}{1}{1} \\
		\ganttbar{Data Screening}{2}{7} \\
		\ganttmilestone{Sharing Data Screening results}{7} \\
		\ganttbar{Risk Classification}{8}{9} \\
		\ganttmilestone{Risk classification results}{9} \\
		\ganttbar{Supervision Planning optimization}{10}{11} \\
		\ganttmilestone{Sharing Final Results}{11} \\
		\ganttbar{Paper Writing}{12}{14} \\
		\ganttmilestone{Paper Submission}{14}\\
	\end{ganttchart}
	\caption{Dissertation Timeline}
\end{figure}
